Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-28 paper.tex

To demonstrate the versatility of \texttt{vcf2zarr}, in this section
we include a case study of a dataset originating from a species whose
genome properties differ substantially from human. The conifer Norway
spruce, \emph{Picea abies}, is one of the largest and most
ecologically important species on Earth and is distributed over large
parts of the Northern hemisphere. One of its features is its large
genome, consisting of 12 autosomes with a genome size of 19.6
Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
resequenced spruce individuals from a recent study (in preparation),
where resequencing data has been mapped to a chromosome-scale
reference genome. The dataset consists of approximately 3.8 million
single-nucleotide variants and small indels from the 12 autosomes
totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
distributed over 165 VCF files. Since the dataset originally was
generated for downstream analyses that require genotype likelihoods,
the genotypes include the PL field.

Following the conversion process detailed in the previous section, we
first converted the 165 VCF files (7.33TiB) to ICF, followed by
conversion to Zarr. The tasks were run as single jobs on a compute
node with 512 GB RAM and 128 cores using 120 cores. Conversion to ICF
required 13h45min runtime. The ICF representation used a total of 6.77
TiB over 1,340,804 data storage files. Conversion to Zarr required
18h36min runtime, generating a dataset with 32 arrays, consuming a
total of 6.6 TiB storage over 10,861,038 chunk files. The reduction in
total storage (1.1X) is negligible, mainly due to the inherent
difficulty in compressing the PL field. The output dataset storage
size is dominated by a major component, call\_PL, which accounts for
nearly 93\% of the total. call\_PL is yet another field with low
compression potential due to inherent noisiness.

Therefore, to reduce storage requirements we generated a Zarr schema
for encoding that uses local allele fields with \texttt{vcf2zarr
mkschema --local-alleles}, and used the generated schema as input to
\texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
generating a dataset with 35 arrays, consuming a total of 2.3TiB
storage over 11,235,556 chunk files, with a reduction in storage of
3.2X. Therefore, by using local allele fields we achieve approximately
3X improvement of compression. The top fields in terms of storage are
detailed inTable~\ref{tab-spruce-data}.

@@

Large genomes from non-model organisms pose particular problems for
analysis workflows that often are tailored for human conditions. For
instance, the binary Sequence Alignment/Map index format
BAI~\cite{li2009sequence} has a 512
Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
The spruce BAM files used for variant calling were BAI indexed, which
motivated the partitioning into multiple VCF files in the first place,
adding complexity to downstream processing.

A similar limitation is imposed by the VCF POS field which is encoded
in 32-bit format and can hold chromomosomes up to 2.14 Gbp. Although
spruce does not have chromosomes that exceed the limit, other species
do, such as the 87.2 Gbp South American lungfish, whose 19 chromosomes
are all but one larger than the 3.1 Gbp human
genome~\cite{schartl_GenomesAllLungfish_2024}. Zarr's flexible type
system means that coordinates can be stored as 64 or 32 bit integers
as needed, and there is therefore no particular limit on genome size.
% limitation, and the coordinate system could easily be converted to
% true genome coordinates for storage in Zarr. This updating procedure
% could be applied to any species of choice.

>>>>

To demonstrate the versatility of \texttt{vcf2zarr}, in this section
we include a case study of a dataset originating from a species whose
genome properties differ substantially from human. The conifer Norway
spruce, \emph{Picea abies}, is one of the largest and most
ecologically important species on Earth and is distributed over large
parts of the Northern hemisphere. One of its features is its large
genome, consisting of 12 autosomes with a genome size of 19.6
Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
resequenced spruce individuals from a recent study (in preparation),
where resequencing data has been mapped to a chromosome-scale
reference genome. The dataset consists of approximately 3.8 million
single-nucleotide variants and small indels from the 12 autosomes
totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
distributed over 165 VCF files. 
% Since the dataset originally was
% generated for downstream analyses that require genotype likelihoods,
% the genotypes include the PL field.

We first converted the 165 VCF files (7.33TiB) to ICF, followed by
conversion to Zarr. The tasks were run as single jobs on a compute
node with 512 GB RAM and 128 cores using 120 cores. Conversion to ICF
required 13h45min runtime. The ICF representation used a total of 6.77
TiB over 1,340,804 data storage files. Conversion to Zarr 
using the default schema and a chunk size of 10,000 variants
$\times$ 1,063 samples
required 18h36min runtime, generating a dataset with 34 arrays, 
and required a total of 6.6 TiB storage over 10,861,038 chunk files. 
Over 90\% of this was used by the call\_PL field, mostly due to 
the maximum number of alleles at a site being 4,
and consequently the inner dimension of call\_PL 10 because 
of its quadratic dependency on the number of alleles (see Methods).

We therefore generated a vcf2zarr schema
using local allele fields with \texttt{vcf2zarr
mkschema --local-alleles}, and used the generated schema as input to
\texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
generating a dataset with 35 arrays, consuming a total of 2.3TiB
storage over 11,235,556 chunk files. 
Using local allele fields in this example we therefore 
achieve approximately 3X improvement of compression. 
The top fields in terms of storage are
detailed in Table~\ref{tab-spruce-data}.

@@

Large genomes from non-model organisms pose particular problems for
analysis workflows that often are tailored for human conditions. For
instance, the binary Sequence Alignment/Map index format
BAI~\cite{li2009sequence} has a 512
Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
The spruce BAM files used for variant calling were BAI indexed, which
motivated the partitioning into multiple VCF files in the first place,
adding complexity to downstream processing.
A similar limitation is imposed by the VCF POS field which is encoded
in 32-bit format and can hold chromomosomes up to 2.14 Gbp. Although
spruce does not have chromosomes that exceed the limit, other species
do, such as the 87.2 Gbp South American lungfish, whose 19 chromosomes
are all but one larger than the 3.1 Gbp human
genome~\cite{schartl_GenomesAllLungfish_2024}. Zarr's flexible type
system means that coordinates can be stored as 64 or 32 bit integers
as needed, and there is therefore no particular limit on genome size.
% limitation, and the coordinate system could easily be converted to
% true genome coordinates for storage in Zarr. This updating procedure
% could be applied to any species of choice.
